---
title: "Lab 18: Pertussis Mini-Project"
author: "Bernice Lozada (A16297973)"
format: pdf
---

Pertussis (whooping cough) is a highly contagious lung infection that is most deadly for people under 1 year of age. 

First, let's look at Pertussis case numbers per year in the US.

The CDC tracks Pertussis case numbers and makes the data available here: 
https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html 

> Question 1

```{r}
cdc <- data.frame(year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )
```

I want a plot of case number per year.

```{r}
library(ggplot2)
base <- ggplot(cdc) + aes(x=year, y=cases) + geom_line() + geom_point()
```

> Q2. Using the ggplot `geom_vline()` function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
base + geom_vline(xintercept = 1946, col = "blue") + geom_vline(xintercept = 1996, col = "red") + geom_vline(xintercept=2020, col = "purple")
```
After the 1946 introduction of the wP vaccine (blue line), the case numbers dropped significantly, meaning the vaccine was highly effective.

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine (red line), the cases started to rise. One reason could be less adherence to vaccines (not getting the booster vs. the one shot needed for the wP vaccine) and being less willing to get vaccinated after the lawsuits over the wP vaccine. 

## CMI-PB

A systems vaccinology project to figure out what's going on with aP vs wP immune responses.

The resource has an API (application programming interface) that returns JSON format data.

Basically "key":"value" pair format.

We will use the jsonlite package to read this data into R.
```{r}
# Allows us to read, write and process JSON data
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 

head(subject, 3)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

There are 60 aP vaccinated subjects and 58 wP vaccinated subjects in the dataset.

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```
There are 79 female and 39 male subjects in the dataset.

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
# read specimen and ab-titer tables into R
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = TRUE) 
```

I need to merge (join) these tables to get all the meta data I need about subjects and specimens in one place. We will use **dplyr** `join()` functions for this task. 

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)

head(meta)
```

Now we can join our `meta` table and join it with our Ab table `ab_titer`.

```{r}
abdata <- inner_join(titer, meta)
head(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype? What ab are measured/recorded in the `titer` table?

```{r}
table(titer$isotype)
```

```{r}
table(titer$antigen)
```
We have our merged dataset with all the needed metadata and antibody measurements called `abdata`.


```{r}
head(abdata,2)
```

## Examine IgG Ab titer levels

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)

```

```{r}
ggplot(igg) + aes(x = MFI, y = antigen) + geom_boxplot() + facet_wrap(vars(visit), nrow=2)
```

```{r}
ggplot(igg) + aes(x = MFI_normalised, y = antigen, col=igg$infancy_vac) + geom_boxplot() +xlim(0,75)+facet_wrap(vars(visit), nrow=2)
```

## Focusing on 2021 dataset IgG PT antigen

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") + labs(title = "2021 dataset - IgG PT")

```

